They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-20 paper/paper.tex
working thru p_ms

Studying both player streaks and team streaks in basketball, in both
observational and controlled settings. Gilovich et al. (1985),
henceforth GVT, do not find evidence for the hot hand phenomenon.
However, Miller and Sanjurjo (2018), henceforth MS, refuted the
conclusions of GVT, proving a framework for quantifying streak selection
bias, which effectively works in the manner posited by the gambler's
fallacy\footnote{Streak selection bias, ``in conjunction with a
  quasi-Bayesian model of decision making under sample size
  neglect\ldots{} provides a novel structural candidate explanation for
  the persistence of gambler's fallacy beliefs'', per MS (2018).}.
Specifically, MS say that a ``bias exists in a common measure of the
conditional dependence of present outcomes on streaks of past outcomes
in sequential data'' implying that, under i.i.d. conditions, ``the
proportion of successes among the outcomes that immediately follow a
streak of consecutive successes is expected to be strictly less than the
underlying (conditional) probability of success''. When applying streak
selection bias to GVT.'s data, MS come to the opposite conclusions as
GVT.

@@

@@

When considering momentum in consecutive rounds wins, I use the
descriptor ``hindsight'' because I account for the length of the series
(\(r\) rounds). Without accounting for the streak selection bias
described by MS (2018), the ``hindsight'' Bernoulli round win
probability immediately following a streak of \(k\) round wins for a
best-of-\(s\) series, \(p^{w|kr}\), is

\begin{equation}\protect\hypertarget{eq:pwkr}{}{
  p^{w|kr} = p(\text{win | k, r})=\{
    \begin{array}{cl}
    \frac{m - k}{i - k}, & \text{team wins series} \\
    \frac{s - m - k}{i - k}, & \text{team loses series}
    \end{array}.
  }
\label{eq:pwkr}
\end{equation}

The hindsight Bernoulli round loss probability after a streak of \(k\)
round wins, \(p^{l|kr}\), can be formulated symmetrically.

However, as shown by MS, we should expect the proportion of rounds wins
immediately following a streak \(k\) rounds wins for a series lasting
\(r\) rounds, \(\hat{P}^{w|kr}\), to be strictly less than
\(p^{w|kr}\).\footnote{See Appendix E.1 of MS (2018) for details on the
  proof that follows.}

\begin{quote}
Theorem 1: \emph{Let}
\(\mathbf{X} = \{X_i\}^{n}_{i=1}, n \geq 3\)\emph{, be a sequence of
independent Bernoulli trials, each with probability of success}
\(0 < p < 1\)\emph{. Let} \(\hat{P}_k(\mathbf{X})\) \emph{be the
proportion of successes on the subset of trials} \(I_k(\mathbf{X})\)
\emph{that immediately follow} \(k\) consecutive successes, that is,
\(\hat{P}_k(\mathbf{X}) := \sum_{i \in I_k(\mathbf{X})} X_i | I_k(\mathbf{X}) |. \hat{P}_k\)
\emph{is a biased estimate of}
\(\mathbb{P}(X_t = 1 | \prod_{j=t-k}^{t-1} X_j = 1) \equiv p\) \emph{for
all} \(k\) \emph{such that} \(1 \leq k \leq n - 2\)\emph{. In
particular,}
\(E[\hat{P}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset] < p.\)
\end{quote}

MS (2018) note that there does not exists a closed form representation
for \(E[\hat{P}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]\)
(\(\hat{P}_k(\mathbf{X})\) is the analogue for \(\hat{P}^{w|kr}\) in my
CoD SnD notation), for \(k > 1\). Figure 1 shows how the expected value
varies as a function of the total number of trials \(n\), given \(p\)
and \(k > 1\).\footnote{I've adapted the code from Vafa (2017), which
  implements MS's framework (2018).}

\begin{figure}
\centering
\includegraphics{images/npkd.png}
\caption{The expected value of the proportion of successes on trials that immediately follow $k$ consecutive successes, $E[\hat{P}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]$, for Bernoulli trial success $p = 0.5$ and streak lengths $k \in [2, 3, 4]$, as a function of the total number of trials $n.$}
\end{figure}

@@

Given that people typically perceive streaks as beginning after the
third success (or failure) at minimum (Carlson and Shu 2007), I focus on
streaks of three round wins.\footnote{Three happens to also be a
  reasonable number for series that lasts at maximum 11 rounds.} Table
\ref{tbl:cod-round-win-prop-after-k-wins} shows \(P^{w,kr}\) and
\(P^{l,kr}\) for round win rate and loss rate following \(k = 3\) round
win streaks respectively.

\begin{longtable}{rrrrrrr}
\caption{First, observed proportion of wins ($\hat{P}^{kr,\text{win}}_O$) and expected proportion of wins, both un-adjusted ($\hat{P}^{kr,\text{win}}_{E,\text{unadj.}}$) and adjusted ($\hat{P}^{kr,\text{win}}_{E,\text{adj.}}$), after 3-round win streak, for a series lasting $r$ rounds. Second, observed proportion of losses ($\hat{P}^{kr,\text{loss}}_O$) and expected proportion of losses, both un-adjusted ($\hat{P}^{kr,\text{loss}}_{E,\text{unadj.}}$) and adjusted ($\hat{P}^{kr,\text{loss}}_{E,\text{adj.}}$), after 3-round loss streak}\label{tbl:cod-round-win-prop-after-k-wins} \\
\toprule
\multicolumn{7}{c}{\text{3 round win streak}} \\
$r$ & $N^{kr,\text{win}}_O$ & $\hat{P}^{kr,\text{win}}_O$ & $\hat{P}^{kr,\text{win}}_{E,\text{unadj.}}$ & $\hat{P}^{kr,\text{win}}_{E,\text{adj.}}$ \\ 
\midrule
7 & 209 & $75\%$ & $75\%$ (0.58) & $61\%$ (<=0.01) \\
8 & 209 & $62\%$ & $60\%$ (0.28) & $45\%$ (<=0.01) \\
9 & 193 & $52\%$ & $50\%$ (0.33) & $35\%$ (<=0.01) \\
10 & 151 & $44\%$ & $43\%$ (0.45) & $29\%$ (<=0.01) \\
11 & 150 & $40\%$ & $38\%$ (0.29) & $25\%$ (<=0.01) \\
\toprule
\multicolumn{7}{c}{\text{3 round loss streak}} \\
$r$ & $N^{kr,\text{loss}}_O$ & $\hat{P}^{kr,\text{loss}}_O$ & $\hat{P}^{kr,\text{loss}}_{E,\text{unadj.}}$ & $\hat{P}^{kr,\text{loss}}_{E,\text{adj.}}$ \\ 
\midrule
10 & 60 & $13\%$ & $14\%$ (0.64) & $10\%$ (0.26) \\ 
11 & 129 & $24\%$ & $25\%$ (0.63) & $16\%$ (<=0.01) \\ 
\bottomrule
\end{longtable}

@@

\hypertarget{appendix}{%
\section*{Appendix}\label{appendix}}
\addcontentsline{toc}{section}{Appendix}

>>>>

Studying both player streaks and team streaks in basketball, in both
observational and controlled settings. Gilovich et al. (1985),
henceforth GVT, do not find evidence for the hot hand phenomenon.
However, Miller and Sanjurjo (2018), henceforth MS, refuted the
conclusions of GVT, proving a framework for quantifying streak selection
bias, which effectively works in the manner posited by the gambler's
fallacy\footnote{Streak selection bias, ``in conjunction with a
  quasi-Bayesian model of decision making under sample size
  neglect\ldots{} provides a novel structural candidate explanation for
  the persistence of gambler's fallacy beliefs'', per MS.}.
Specifically, MS say that a ``bias exists in a common measure of the
conditional dependence of present outcomes on streaks of past outcomes
in sequential data'' implying that, under i.i.d. conditions, ``the
proportion of successes among the outcomes that immediately follow a
streak of consecutive successes is expected to be strictly less than the
underlying (conditional) probability of success''. When applying streak
selection bias to GVT.'s data, MS come to the opposite conclusions as
GVT.

@@

\begin{equation}\protect\hypertarget{eq:m}{}{
m = \frac{s + 1}{2}
}\label{eq:m}\end{equation}

@@

Removing our knowledge of a streak, we might model the Bernoulli round
win probability as

\begin{equation}\protect\hypertarget{eq:pwr}{}{
p^{w|r} = p(\text{win | r})=\{
\begin{array}{cl}
\frac{m}{r}, & \text{team wins series} \\
\frac{r - m}{r}, & \text{team loses series}
\end{array}.
}
\label{eq:pwr}
\end{equation}

The Bernoulli round loss probability after a streak of \(k\) round wins,
\(p^{\ell|kr}_0\), can be formulated symmetrically.

As shown by MS, we should expect the proportion of rounds wins
immediately following a streak of \(k\) rounds wins for a series lasting
\(r\) rounds, \(\hat{P}^{w|kr}_{MS}\), to be strictly less than
\(p^{w|r}\).\footnote{See Appendix E.1 of MS for details on the proof
  that follows.}

\begin{quote}
Theorem 1: \emph{Let}
\(\mathbf{X} = \{X_i\}^{n}_{i=1}, n \geq 3\)\emph{, be a sequence of
independent Bernoulli trials, each with probability of success}
\(0 < \pi < 1\)\emph{. Let} \(\hat{\Pi}_k(\mathbf{X})\) \emph{be the
proportion of successes on the subset of trials} \(I_k(\mathbf{X})\)
\emph{that immediately follow} \(k\) consecutive successes, that is,
\(\hat{\Pi}_k(\mathbf{X}) := \sum_{i \in I_k(\mathbf{X})} X_i | I_k(\mathbf{X}) |. \hat{\Pi}_k\)
\emph{is a biased estimate of}
\(\mathbb{\Pi}(X_t = 1 | \prod_{j=t-k}^{t-1} X_j = 1) \equiv p\)
\emph{for all} \(k\) \emph{such that} \(1 \leq k \leq n - 2\)\emph{. In
particular,}
\(E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset] < \pi.\)
\end{quote}

MS note that there does not exists a closed form representation for
\(E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]\) for
\(k > 1\). (\(\hat{\Pi}_k(\mathbf{X})\) is the analogue for
\(\hat{P}^{w|kr}_{MS}\) in my notation for CoD SnD.) Figure 1 shows how
the expected value varies as a function of the total number of trials
\(n\), given \(\pi\) and \(k > 1\).\footnote{I've adapted the code from
  Vafa (2017), which implements MS's framework (2018).}

\begin{figure}
\centering
\includegraphics{images/pwkr_ms.png}
\caption{The expected value of the proportion of successes on trials that immediately follow $k$ consecutive successes, $E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]$, for Bernoulli trial success $\pi = 0.5$ and streak lengths $k \in [2, 3, 4]$, as a function of the total number of trials $n.$}
\end{figure}

My context is fundamentally different from that of GVT and MS, both of
whom focus on longitudinal data in controlled settings. GVT also perform
statistical tests on shots from players in live games,
i.e.~``observational'' data, but they note that their findings are
likely affected player shot selection in the face of defensive strategy
by the opposing team. The number of trials is fixed in their
experimental designs, but in CoD SnD, the number of rounds played is
determined as a function of the max number of possible rounds (\(s\))
and whether or not the team wins the series, as shown in Equation
\ref{eq:pwr}. The Bernoulli trial success probability, i.e.~a round win
in CoD SnD, is not independent of the opponent.

Consequently, a statistical test of the difference in \(\hat{P}^{w|kr}\)
and \(\hat{P}^{\ell|kr}\), as performed by GVT and MS to evaluate their
respective hypothesis regarding
\(E[\hat{P}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]\) (Theorem
1), is not appropriate.

Let us consider another form of the expected proportion of rounds won
immediately after a streak of \(k\) round wins in a best-of-\(s\) series
given the length of the series (\(r\) rounds), the ``hindsight''
proportion \(\hat{P}^{w|kr}_0\). The Bernoulli round win probability
underlying \(\hat{P}^{w|kr}_0\) is

\begin{equation}\protect\hypertarget{eq:pwkr}{}{
p_0^{w|kr} = p_0(\text{win | k, r})=\{
\begin{array}{cl}
\frac{m - k}{i - k}, & \text{team wins series} \\
\frac{s - m - k}{i - k}, & \text{team loses series}
\end{array}.
}
\label{eq:pwkr}
\end{equation}

A binomial test for the expected proportion of round wins following
streaks of \(k\) round wins in a series lasting \(r\) rounds,
\(\hat{P}^{w|kr}_0\), assuming \(p^{w|kr}_0\) as the probability of
success, can be used to verify that \(\hat{P}^{w|kr}_0\) is reasonable
as to use as a baseline with which to compare \(\hat{P}^{w|kr}_{MS}\).
If the binomial test shows that we cannot reject the null hypothesis
that the expected \(\hat{P}^{w|kr}_0\) is different from the observed
proportion \(P^{w|kr}\).

A proportion test of \(\hat{P}^{w|kr}_0\) and \(\hat{P}^{w|kr}_{MS}\)
can tell us whether the momentum exists in CoD SnD. A null result means
we cannot say that there is a difference between the hindsight expected
proportion \(\hat{P}^{w|kr}_0\) and the expected proportion adjusted for
streak-selection bias, \(\hat{P}^{w|kr}_0\).

@@

Given that people typically perceive streaks as beginning after the
third success (or failure) at minimum (Carlson and Shu 2007), I focus on
streaks of three round wins.\footnote{Three happens to also be a
  reasonable number for series that lasts at maximum 11 rounds.} Table
\ref{tbl:cod-pw3r-pl3r} shows \(P^{w,kr}\) and \(P^{l,kr}\) for round
win rate and loss rate following \(k = 3\) round win streaks
respectively.

First,

\begin{longtable}{rrrrrrr}
\caption{Observed count and proportion of round wins immediately following round win streak of $k=3$, $N^{w|kr}$ and $P^{w|kr}$ respectively, in series lasting $r$ rounds. Additionally, notional expected proportion, $\hat{P}^{w|kr}_0$ and binomial test p-value, assuming notional round probability $p^{w|kr}$ and using observed counts; MS expected proportion $\hat{P}^{w|kr}_{MS}$; and $\hat{D}^{w|kr} = \hat{P}^{w|kr}_{MS} - \hat{P}^{w|kr}_0$ and proportion test p-value. Fields presented in groups, split by whether the team won the round immediately following the 3-round win streak, and whether the team won the series. The post-streak loss analogues are presented as $N^{\ell|kr}$, $P^{\ell|kr}$, $\hat{P}^{\ell|kr}_0$, $\hat{P}^{\ell|kr}_{MS}$, and $\hat{D}^{\ell|kr}$.}
\label{tbl:cod-pw3r-pl3r} \\
\toprule
\multicolumn{6}{c}{\text{win series, win after 3-round win streak}} \\
\midrule
$r$ & $N^{w|kr}$ & $P^{w|kr}$ & $\hat{P}^{w|kr}_0$ ($p$-value) & $\hat{P}^{w|kr}_{MS}$ & $\hat{D}^{w|kr}$ ($p$-value) \\ 
\midrule
7 & 209 & 74.6\% & 75.0\% (0.58) & 75.7\% & -0.7\% (0.94) \\ 
8 & 209 & 62.2\% & 60.0\% (0.28) & 61.9\% & -1.9\% (1.00) \\ 
9 & 193 & 51.8\% & 50.0\% (0.33) & 52.7\% & -2.7\% (0.97) \\ 
10 & 151 & 43.7\% & 42.9\% (0.45) & 44.7\% & -1.8\% (0.94) \\ 
11 & 150 & 40.0\% & 37.5\% (0.29) & 38.8\% & -1.3\% (1.00) \\ 
\toprule
\toprule
\multicolumn{6}{c}{\text{lose series, win after 3-round win streak}} \\
\midrule
$r$ & $N^{w|kr}$ & $P^{w|kr}$ & $\hat{P}^{w|kr}_0$ ($p$-value) & $\hat{P}^{w|kr}_{MS}$ & $\hat{D}^{w|kr}$ ($p$-value) \\ 
\midrule
10 & 60 & 13.3\% & 14.3\% (0.64) & 26.7\% & -12.4\% (1.00) \\ 
11 & 129 & 24.0\% & 25.0\% (0.63) & 30.8\% & -5.8\% (0.98) \\ 
\toprule
\toprule
\multicolumn{6}{c}{\text{win series, lose after 3-round win streak}} \\
\midrule
$r$ & $N^{\ell|kr}$ & $P^{\ell|kr}$ & $\hat{P}^{\ell|kr}_0$ ($p$-value) & $\hat{P}^{\ell|kr}_{MS}$ & $\hat{D}^{\ell|kr}$ ($p$-value) \\ 
\midrule
7 & 209 & 25.4\% & 25.0\% (0.48) & 6.4\% & 18.6\% (0.98) \\ 
8 & 209 & 37.8\% & 40.0\% (0.76) & 14.7\% & 25.3\% (0.96) \\ 
9 & 193 & 48.2\% & 50.0\% (0.72) & 22.2\% & 27.8\% (0.97) \\ 
10 & 151 & 56.3\% & 57.1\% (0.62) & 26.7\% & 30.5\% (0.99) \\ 
11 & 150 & 60.0\% & 62.5\% (0.76) & 30.8\% & 31.7\% (0.95) \\ 
\toprule
\toprule
\multicolumn{6}{c}{\text{lose series, lose after 3-round win streak}} \\
\midrule
$r$ & $N^{\ell|kr}$ & $P^{\ell|kr}$ & $\hat{P}^{\ell|kr}_0$ ($p$-value) & $\hat{P}^{\ell|kr}_{MS}$ & $\hat{D}^{\ell|kr}$ ($p$-value) \\ 
\midrule
10 & 60 & 86.7\% & 85.7\% (0.51) & 44.7\% & 41.1\% (1.00) \\ 
11 & 129 & 76.0\% & 75.0\% (0.45) & 38.8\% & 36.2\% (0.95) \\ 
\bottomrule
\end{longtable}

@@
